\pkg{cpop}: Detecting changes in piecewise-linear signals

### 2022-08-05 latex/article.tex
Add files via upload

%% almost as usual
\author{Paul Fearnhead\\Lancaster Universtity \And 
        Dan Grose\\Lancaster University}
\title{\pkg{cpop}: Detecting changes in piecewise-linear signals}

@@

The following code demonstrates the \code{simulate} function and displays the data along with the (true) line segments and the locations of the changes in slope (see Figure \ref{fig:simulate-example}).
\begin{CodeChunk}
\begin{CodeInput}
R> library("cpop")
R> library("ggplot2")
R> changepoints <- c(0, 25, 50, 100)
R> change.slope <- c(0.2, -0.3, 0.2, -0.1)
R> x <- 1:200
R> sd <- 0.8
R> y <- simulate(x, changepoints, change.slope, sd)
R> df <- data.frame("x" = x, "y" = y)
R> p <- ggplot(data = df, aes(x = x, y = y))
R> p <- p + geom_point()
R> p <- p + geom_vline(xintercept = changepoints,
+                      color = "red",
+                      linetype = "dashed")
R> mu <- simulate(x, changepoints, change.slope, sd = 0)
R> p <- p + geom_line(aes(y = mu), color = "blue")
R> p <- p + theme_bw()
R> print(p)
\end{CodeInput}
\end{CodeChunk}
%
%
\begin{figure}
\centering
%\includegraphics[scale=0.5]{figures/simulate-example.pdf}
\includegraphics[width=0.6\linewidth]{figures/simulate_example_ggplot.pdf}
\caption{Simulated data (black dots) with true mean (blue dashed line) and changepoints (vertical red dashed lines).}
\label{fig:simulate-example}
\end{figure}
%
%
\subsection{Determining Changes in Slope}
%
%
The function \code{cpop} is used to determine the locations of changes in slope.
%
%
\begin{CodeInput}
cpop(y, x = 1:length(y) - 1, grid = x, beta = 2 * log(length(y)), 
    sd = 1, minseglen = 0, prune.approx = FALSE)
\end{CodeInput}
It takes the following arguments :
%
%
\begin{itemize}
 \item  \code{y} - A vector of length $n$ containing the data.
\item \code{x} -  A vector of length $n$ containing the times/locations of data points. Default value is NULL, in which case the locations are set to be $0,1,\ldots,n-1$, corresponding to evenly spaced data.
\item \code{grid} -  An ordered vector of possible locations for the change points. If this is NULL, then this is set to $x$, the vector of times/locations of the data points.
\item \code{beta} -  A positive real value for the penalty, $\beta$ in (\ref{eq:penalisedcost2}), incurred for adding a changepoint. The larger the penalty, the fewer changepoints will be detected.
\item \code{sd} - Estimate of residual standard deviation. Can be a single numerical value if it is the same for all data points, or a vector of $n$ values for the case of varying standard deviation. 
\item \code{minseglen} -  The minimum allowable segment length, i.e. distance between successive changepoints. The default is that no minimum segment length is imposed.
\item \code{prune.approx} -  Only relevant if a minimum segment length is set. If True, \code{cpop} will use an approximate pruning algorithm that will speed up computation but may occasionally lead to a sub-optimal solution in terms of the estimated changepoint locations. If the minimum segment length is 0, then an exact pruning algorithm is possible and is used.
\end{itemize}
The \code{cpop} function returns an S4 object for which a number of generic methods, including \code{plot} and \code{summary}, are provided. 
%
%
%\subsubsection{simulated data}
The following demonstrates how the \code{cpop} function can be used to determine changes in slope, by analysing the data we simulated and plotted in Figure \ref{fig:simulate-example}. It uses the default penalty, $\beta=2\log n$, and we assume that the true noise standard deviation, 0.8, is known. The \code{summary} function is used to provide an overview of the analysis parameters along with estimated changepoint locations, corresponding fitted line segments, and the (weighted) residual sum of squares (RSS) for each segment. 
%
%
\begin{CodeChunk}
\begin{CodeInput}
R> res <- cpop(y, x, sd = 0.8)
R> summary(res)
\end{CodeInput}
\begin{CodeOutput}
cpop analysis with n = 200 and penalty (beta)  = 10.59663

@@

\begin{CodeChunk}
\begin{CodeInput}
R> x <- (1:200)^(2)/(200)

R> changepoints <- c(0, 25, 50, 100)
R> change.slope <- c(0.2, -0.3, 0.2, -0.1)
R> sd <- 0.8
R> y <- simulate(x, changepoints, change.slope, sd)
\end{CodeInput}
\end{CodeChunk}

To analyse the data we use \code{cpop} as before. (The only difference is that for evenly spaced data one can omit the \code{x} argument -- but it must be included for unevenly spaced data.)
\begin{CodeChunk}
\begin{CodeInput}
R> res <- cpop(y, x, beta = 2 * log( length(y) ) )
\end{CodeInput}
\end{CodeChunk}
%
%
Figure \ref{fig:cpop-example-uneven} shows a plot of the simulated data and estimated changepoints and mean function.

@@

\begin{figure}
\begin{minipage}{.5\linewidth}
\centering
%{\includegraphics[scale=.5]{figures/cpop-example-uneven2.pdf}}
{\includegraphics[scale=.5]{figures/cpop_example_uneven_1_ggplot.pdf}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
%\includegraphics{figures/cpop-example-uneven.pdf}
{\includegraphics[scale=.5]{figures/cpop_example_uneven_2_ggplot.pdf}}
\end{minipage}%
\caption{Example output of \code{cpop} for heterogenous noise: assuming a constant noise variance (left) and the true noise variance (right). The true mean and changepoints are given in blue dashed lines, together with estimated mean (black full line) and changepoints (red full lines). }
\label{fig:cpop-example-heterogeneous}
\end{figure}

@@

We simulated data as per Section \ref{sec:cpop} but with \code{sd = 1.5}. To run CPOP for a range of $\beta\in[7.5, 50]$ we use
\begin{CodeInput}
R> res.crops <- cpop.crops(y , x, beta_min= 7.5 ,beta_max= 50)
\end{CodeInput}
For our example $2 \log n = 10.6$, so this is equivalent to trying noise standard deviation in the range $[0.8,2.2]$.

We can plot the location in the changepoints for each segmentation found by CPOP for $\beta\in[7.5,50]$
\begin{CodeInput}
R> plot(rec.crops)
\end{CodeInput}
This is shown in Figure \ref{fig:crops.cpopA}, and shows that there are 6 different segmentations found. These are labelled with a penalty value which gives that segmentation (left axis) and the unpenalised cost, i.e. the weighted RSS, for that segmentation and penalty value (right axis).

Details of the segmentations can be obtained using \code{segmentations(res.crops) }. This gives a matrix with one row for each of the segmentations, and each row contains a value of $\beta$ that gives that segmentation, the corresponding unpenalised cost, the penalised cost, the number of changepoints, and then the list of ordered changepoints. 
%
%
We can also obtain a list with the output from \code{cpop} corresponding to each segmentation, with \code{models( res.crops )}. For example, one approach to choose a segmentation from the output from \code{cpop.crops} is to find the segmentation that minimises a cost under a model where we assume the noise variance is unknown \citep{fryzlewicz2014wild},
\[
 n \log \left( \frac{1}{n} \sum_{i=1}^n \left(y_i-\hat{f}(x_i)\right)^2
\right) + 2 K \log n.
\]
Here $\hat{f}$ is the estimated mean function and $K$ is the number of changepoints. This can be calculated as follows.
\begin{CodeChunk}
\begin{CodeInput}
R> models <- cpop.crops.models(res.crops)
R> M <- length(models)
R> BIC <- rep(NA, M)
R> K <- segmentations( res.crops )[,4]
R> n <- length(y)
R> for(j in 1:M){
R>   BIC[i] <- n * log( mean( ( residuals(models[[j]]) )^2) )
   + 2 * K[i]  * log(n)
R> }
\end{CodeInput}
\end{CodeChunk}
This uses that the fourth column of the matrix \code{segmentations( res.crops )} stores the number of changepoints in each segmentation, and that we can calculate $y_i - \hat{f}(x_i)$ using the \code{residual} function evaluated for the corresponding entry of \code{cpop.crops.models(res.crops)}. The segmentation which has the smallest value of \code{BIC} is shown in Figure \ref{fig:crops.cpopA}, and shows that this correctly chooses the segmentation with three changes.

\begin{figure}
\begin{minipage}{.5\linewidth}
\centering
{\includegraphics[scale=.5]{figures/cpop-cropsA1.pdf}}
\end{minipage}%
\begin{minipage}{.5\linewidth}
\centering
{\includegraphics[scale=.5]{figures/cpop-cropsA2.pdf}}
\end{minipage}%
\caption{Example plot of output from \code{cpop.crops} (left), and best segmentation based on calculated \code{BIC} (right). For the left-hand plot each row shows a segmentation, with points at estimated changepoint location. The left-hand axis shows a penalty value that leads to that segmentation, and the right-hand axis gives the corresponding unpenalised cost. For the right-hand plot we show the true mean and changepoints (blue dashed lines), and estimated mean (black line) and changepoints (red lines).}
\label{fig:crops.cpopA}
\end{figure}

As a final example, we performed a similar analysis but with correlated noise. This violates the assumption of IID noise that underpins the default choice of penalty, thus we run \code{cpop.crops} for a range of penalties.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> res.crops <- cpop.crops(y,x,beta_min=8,beta_max=200)
R> segs <- segmentations(res.crops)
R> plot(segs[,"m"], segs[,"Qm"], type="l", 
+  xlab="No. of changepoints" ,ylab="Unpenalised Cost")
\end{CodeInput}
\end{CodeChunk}
This runs \code{cpop.crops} and then uses the fact that the output of \code{segmentations} includes columns that give the number of changepoints and the unpenalised cost of each segmentation. These columns are labelled \code{"m"} and \code{"Qm"} respectively.
The plot gives a clear elbow, see Figure \ref{fig:crops.cpopB}, and this corresponds to a correct estimate of the number of changes. %The fitted changes and estimated mean function are also shown in Figure \ref{fig:crops.cpopB}.
%
%\begin{figure}
%\begin{minipage}{.5\linewidth}
%\centering
%{\includegraphics[scale=.5]{figures/cpop-cropsB1.pdf}}
%\end{minipage}%
%\begin{minipage}{.5\linewidth}
%\centering
%{\includegraphics[scale=.5]{figures/cpop-cropsB2.pdf}}
%\end{minipage}%
%\newline
%\begin{minipage}{1.0\linewidth}
%\centering
%{\includegraphics[scale=.5]{figures/cpop-cropsB3.pdf}}
%\end{minipage}%
%\caption{Correlated noise example. Output from \code{cpop.crops} (top left), unpenalised cost against number of changepoint (rop right) and estimate from segmentation corresponding to "elbow" (bottom).}
%\label{fig:crops.cpopB}
%\end{figure}
%
\begin{figure}
\centering
{\includegraphics{figures/cpop_crops_ggplot.pdf}}
\caption{Correlated noise example. Output from \code{cpop.crops} (top left), unpenalised cost against number of changepoint (rop right) and estimate from segmentation corresponding to "elbow" (bottom).}
\label{fig:crops.cpopB}
\end{figure}

>>>>

%% almost as usual
\author{Paul Fearnhead\\Lancaster Universtity \And 
        Daniel Grose\\Lancaster University}
\title{\pkg{cpop}: Detecting changes in piecewise-linear signals}

@@

The following code demonstrates the \code{simulate} function and displays the data along with the (true) line segments and the locations of the changes in slope (see Figure \ref{fig:simulate-example}).
\begin{CodeChunk}
\begin{CodeInput}
R> library("cpop")
R> library("ggplot2")
R> changepoints <- c(0, 25, 50, 100)
R> change.slope <- c(0.2, -0.3, 0.2, -0.1)
R> x <- 1:200
R> sd <- 0.8
R> y <- simulate(x, changepoints, change.slope, sd)
R> df <- data.frame("x" = x, "y" = y)
R> p <- ggplot(data = df, aes(x = x, y = y))
R> p <- p + geom_point(alpha = 0.4)
R> p <- p + geom_vline(xintercept = changepoints,
+                      color = "red",
+                      linetype = "dashed")
R> mu <- simulate(x, changepoints, change.slope, sd = 0)
R> p <- p + geom_line(aes(y = mu), color = "blue")
R> p <- p + theme_bw()
R> print(p)
\end{CodeInput}
\end{CodeChunk}
%
%
\begin{figure}
\centering
%\includegraphics[scale=0.5]{figures/simulate-example.pdf}
\includegraphics[width=0.6\linewidth]{figures/simulate_example_ggplot.pdf}
\caption{Simulated data (black dots) with true mean (blue dashed line) and changepoints (vertical red dashed lines).}
\label{fig:simulate-example}
\end{figure}
%
%
\subsection{Determining Changes in Slope}
%
%
The function \code{cpop} is used to determine the locations of changes in slope.
%
%
\begin{CodeInput}
cpop(y, x = 1:length(y) - 1, grid = x, beta = 2 * log(length(y)), 
    sd = 1, minseglen = 0, prune.approx = FALSE)
\end{CodeInput}
It takes the following arguments :
%
%
\begin{itemize}
 \item  \code{y} - A vector of length $n$ containing the data.
\item \code{x} -  A vector of length $n$ containing the times/locations of data points. Default value is NULL, in which case the locations are set to be $0,1,\ldots,n-1$, corresponding to evenly spaced data.
\item \code{grid} -  An ordered vector of possible locations for the change points. If this is NULL, then this is set to $x$, the vector of times/locations of the data points.
\item \code{beta} -  A positive real value for the penalty, $\beta$ in (\ref{eq:penalisedcost2}), incurred for adding a changepoint. The larger the penalty, the fewer changepoints will be detected.
\item \code{sd} - Estimate of residual standard deviation. Can be a single numerical value if it is the same for all data points, or a vector of $n$ values for the case of varying standard deviation. 
\item \code{minseglen} -  The minimum allowable segment length, i.e. distance between successive changepoints. The default is that no minimum segment length is imposed.
\item \code{prune.approx} -  Only relevant if a minimum segment length is set. If True, \code{cpop} will use an approximate pruning algorithm that will speed up computation but may occasionally lead to a sub-optimal solution in terms of the estimated changepoint locations. If the minimum segment length is 0, then an exact pruning algorithm is possible and is used.
\end{itemize}
The \code{cpop} function returns an S4 object for which a number of generic methods, including \code{plot} and \code{summary}, are provided. 
%
%
%\subsubsection{simulated data}
The following demonstrates how the \code{cpop} function can be used to determine changes in slope, by analysing the data we simulated and plotted in Figure \ref{fig:simulate-example}. It uses the default penalty, $\beta=2\log n$, and we assume that the true noise standard deviation, 0.8, is known. The \code{summary} function is used to provide an overview of the analysis parameters along with estimated changepoint locations, corresponding fitted line segments, and the (weighted) residual sum of squares (RSS) for each segment. 
%
%
\begin{CodeChunk}
\begin{CodeInput}
R> res <- cpop(y, x, sd = 0.8)
R> summary(res)
\end{CodeInput}
\begin{CodeOutput}
cpop analysis with n = 200 and penalty (beta)  = 10.59663

@@

\begin{CodeChunk}
\begin{CodeInput}
R> x <- (1:200)^(2)/(200)
R> changepoints <- c(0, 25, 50, 100)
R> change.slope <- c(0.2, -0.3, 0.2, -0.1)
R> sd <- 0.8
R> y <- simulate(x, changepoints, change.slope, sd)
\end{CodeInput}
\end{CodeChunk}

To analyse the data we use \code{cpop} as before. (The only difference is that for evenly spaced data one can omit the \code{x} argument -- but it must be included for unevenly spaced data.)
\begin{CodeChunk}
\begin{CodeInput}
R> res <- cpop(y, x, sd = 0.8)
\end{CodeInput}
\end{CodeChunk}
%
%
Figure \ref{fig:cpop-example-uneven} shows a plot of the simulated data and estimated changepoints and mean function.

@@

%\begin{figure}
%\begin{minipage}{.5\linewidth}
%\centering
%{\includegraphics[scale=.5]{figures/cpop-example-uneven2.pdf}}
%{\includegraphics[scale=.5]{figures/cpop_example_uneven_1_ggplot.pdf}}
%\end{minipage}%
%\begin{minipage}{.5\linewidth}
%\centering
%\includegraphics{figures/cpop-example-uneven.pdf}
%{\includegraphics[scale=.5]{figures/cpop_example_uneven_2_ggplot.pdf}}
%\end{minipage}%
%\caption{Example output of \code{cpop} for heterogenous noise: assuming a constant noise variance (left) and the true noise variance (right). The true mean and changepoints are given in blue dashed lines, together with estimated mean (black full line) and changepoints (red full lines). }
%\label{fig:cpop-example-heterogeneous}
%\end{figure}

\begin{figure}
\centering
{\includegraphics[scale=.5,width=14cm,trim={0 4cm 0 4cm},clip]{figures/cpop_uneven_examples_ggplot.pdf}}
\caption{Example output of \code{cpop} for heterogenous noise: assuming a constant noise variance (left) and the true noise variance (right). The true mean and changepoints are given in blue dashed lines, together with estimated mean (black full line) and changepoints (red full lines). }
\label{fig:cpop-example-heterogeneous}
\end{figure}

@@

We simulated data as per Section \ref{sec:cpop} but with \code{sd = 1.5}. To run CPOP for a range of $\beta\in[7.5, 50]$ we use
\begin{CodeInput}
R> res.crops <- cpop.crops(y , x, beta_min= 5 ,beta_max= 50)
\end{CodeInput}
For our example $2 \log n = 10.6$, so this is equivalent to trying noise standard deviation in the range $[0.8,2.2]$.

We can plot the location in the changepoints for each segmentation found by CPOP for $\beta\in[7.5,50]$
\begin{CodeInput}
R> plot(res.crops)
\end{CodeInput}
This is shown in Figure \ref{fig:crops.cpopA}, and shows that there are 6 different segmentations found. These are labelled with a penalty value which gives that segmentation (left axis) and the unpenalised cost, i.e. the weighted RSS, for that segmentation and penalty value (right axis).

Details of the segmentations can be obtained using \code{segmentations(res.crops) }. This gives a matrix with one row for each of the segmentations, and each row contains a value of $\beta$ that gives that segmentation, the corresponding unpenalised cost, the penalised cost, the number of changepoints, and then the list of ordered changepoints. 
%
%
We can also obtain a list with the output from \code{cpop} corresponding to each segmentation, with \code{models( res.crops )}. For example, one approach to choose a segmentation from the output from \code{cpop.crops} is to find the segmentation that minimises a cost under a model where we assume the noise variance is unknown \citep{fryzlewicz2014wild},
\[
 n \log \left( \frac{1}{n} \sum_{i=1}^n \left(y_i-\hat{f}(x_i)\right)^2
\right) + 2 K \log n.
\]
Here $\hat{f}$ is the estimated mean function and $K$ is the number of changepoints. This can be calculated as follows.
\begin{CodeChunk}
\begin{CodeInput}
R> models <- cpop.crops.models(res.crops)
R> M <- length(models)
R> BIC <- rep(NA, M)
R> ncps <- segmentations( res.crops )[,4]
R> n <- length(y)
R> for(j in 1:M){
R>   BIC[i] <- n * log( mean( ( residuals(models[[j]]) )^2) )
   + 2 * ncps[i]  * log(n)
R> }
\end{CodeInput}
\end{CodeChunk}
This uses that the fourth column of the matrix \code{segmentations( res.crops )} stores the number of changepoints in each segmentation, and that we can calculate $y_i - \hat{f}(x_i)$ using the \code{residual} function evaluated for the corresponding entry of \code{cpop.crops.models(res.crops)}. The segmentation which has the smallest value of \code{BIC} is shown in Figure \ref{fig:crops.cpopA}, and shows that this correctly chooses the segmentation with three changes.

%\begin{figure}
%\begin{minipage}{.5\linewidth}
%\centering
%{\includegraphics[scale=.55]{figures/cpop-cropsA1.pdf}}
%\end{minipage}%
%\begin{minipage}{.5\linewidth}
%\centering
%{\includegraphics[scale=.55]{figures/cpop-cropsA2.pdf}}
%\end{minipage}%
%\caption{Example plot of output from \code{cpop.crops}
%(left), and best segmentation based on calculated \code{BIC} %(right). For the left-hand plot each row shows a %segmentation, with points at estimated changepoint location. %The left-hand axis shows a penalty value that leads to that %segmentation, and the right-hand axis gives the %corresponding unpenalised cost. For the right-hand plot we %show the true mean and changepoints (blue dashed lines), and %estimated mean (black line) and changepoints (red lines).}
%\label{fig:crops.cpopA}
%\end{figure}
%
%
\begin{figure}
\centering
{\includegraphics[scale=.5,width=15cm,trim={0 4.4cm 0 4.4cm},clip]{figures/cpop_crops_example_ggplot.pdf}}
\caption{Example plot of output from \code{cpop.crops}
(left), and best segmentation based on calculated \code{BIC} (right). For the left-hand plot each row shows a segmentation, with points at estimated changepoint location. The left-hand axis shows a penalty value that leads to that segmentation, and the right-hand axis gives the corresponding unpenalised cost. For the right-hand plot we show the true mean and changepoints (blue dashed lines), and estimated mean (black line) and changepoints (red lines).}
\label{fig:crops.cpopA}
\end{figure}
%
%
As a final example, we performed a similar analysis but with correlated noise. This violates the assumption of IID noise that underpins the default choice of penalty, thus we run \code{cpop.crops} for a range of penalties.

@@

\begin{CodeChunk}
\begin{CodeInput}
R> res.crops <- cpop.crops(y, x, beta_min = 8, beta_max = 200)
R> segs <- segmentations(res.crops)
R> p <- ggplot(data = segs, aes(x = m))
R> p <- p + geom_line(aes(y = Qm))
R> p <- p + geom_vline(xintercept = 10, color = "red")
R> p <- p + xlab("No. of changepoints") + ylab("unpenalised cost")
R> plot(p)
\end{CodeInput}
\end{CodeChunk}
This runs \code{cpop.crops} and then uses the fact that the output of \code{segmentations} includes columns that give the number of changepoints and the unpenalised cost of each segmentation. These columns are labelled \code{"m"} and \code{"Qm"} respectively.
The plot gives a clear elbow, see Figure \ref{fig:crops.cpopB}, and this corresponds to a correct estimate of the number of changes. %The fitted changes and estimated mean function are also shown in Figure \ref{fig:crops.cpopB}.
%
%\begin{figure}
%\begin{minipage}{.5\linewidth}
%\centering
%{\includegraphics[scale=.5]{figures/cpop-cropsB1.pdf}}
%\end{minipage}%
%\begin{minipage}{.5\linewidth}
%\centering
%{\includegraphics[scale=.5]{figures/cpop-cropsB2.pdf}}
%\end{minipage}%
%\newline
%\begin{minipage}{1.0\linewidth}
%\centering
%{\includegraphics[scale=.5]{figures/cpop-cropsB3.pdf}}
%\end{minipage}%
%\caption{Correlated noise example. Output from \code{cpop.crops} (top left), unpenalised cost against number of changepoint (rop right) and estimate from segmentation corresponding to "elbow" (bottom).}
%\label{fig:crops.cpopB}
%\end{figure}
%
\begin{figure}
\centering
{\includegraphics{figures/cpop_crops_ggplot.pdf}}
\caption{Correlated noise example. Output from \code{cpop.crops} (top left), unpenalised cost against number of changepoint (rop right) and estimate from segmentation corresponding to "elbow" (bottom).}
\label{fig:crops.cpopB}
\end{figure}
